An algorithm for a spatial mapping using a hexagon tilegram, with application to Australia disease maps

### 2019-01-17 algorithm_radix.Rmd
Introduction draft

---
title: "An algorithm for a spatial mapping using a hexagon tilegram, with application to Australia disease maps"
author:
  - name: Stephanie Kobakian
    url: https://github.com/srkobakian
    affiliation: Queensland University of Technology
date: "`r Sys.Date()`"
output: radix::radix_article
---

@@

Representing the globe in 2D space has presented issues for many years. Methods have been established to visualise geographical space.
We present a solution for alternate visualisations when the intention is to represent the spatial distribution.

@@

- Spatial distributions are not well presented when areas differ in size.
- We want to understand the distribution among population
- We can benefit from making the focus the distribution, not the geography

@@

- Presenting spatial distributions
- Extremely small areas are lost not easily compared at a high level
- Choropleths; unequal space on the display, difficult to represent Australia
- Cartograms; strong distortions misrepresent Australia

@@

- Unusual population settlement and dispersion over the geographical space: Australia
Dense south east, sparse central and west

- Applying alternative maps that allow for dense population areas when cities that are population dense are close together

@@

The algorithm operates on a set of polygons.
There are parameters needed that may be provided, or will be automatically derived.
All necessary functions are exported, with a main function `create_hexmap` used to step through automatically.

@@

The `create_hexmap` functions requires several parameters. If not provided the information may be derived from the polygon set used.

**The following must be provided to `create_hexmap`:**

The set of polygons to be made into a hexagon map:
`shp: a shape file`

The location of a set of polygons to be made into a hexagon map:
`shp_path: character string location of shape file`

A column of unique names or id values for each area
`sf_id: name of a unique column that distinguishes areas`

A data frame of reference points to determine the order of allocation
`focal_points a data frame of reference locations when allocating hexagons, capital cities of Australia are provided as capital_cities`

```{r}
library(tidyverse)
library(sugaRbag)

@@

**The following will be determined within `create_hexmap` if not provided, but must be created for the following example:**

- `projstring	a string to indicate the projection and epsg`
- `epsg	the four character string to indicate the CRS`
- `buffer_dist: a float value for distance in degrees to extend beyond the - geometry provided`
- `hex_size	a float value in degrees for the diameter of the hexagons`
- `filter_dist amount of hexagons around centroid to consider for allocation`
- `width the angle used to filter the grid points around a centroid`

These parameters will be created when they are needed in the example:

@@

<!--
needed later:

- filter distance: *the amount of hexagons around each centroid to consider for allocation, by default this is 10 hexagons*
-->

When utilising individual functions, we recommend the following approach:

@@

The collection of polygons should be considered first.
There are several ways of providing the set of polygons.
The following methods are available via the `read_shape` function.

- To utilise a shape file from a directory on your device, provide the file path for .shp file.
- The `st_read` function is applied internally to create the sf object.
- If the shape file is an .Rda file, it is loaded into the current environment.

Empty geometries are removed from the sf object, and the remaining polygons are projected to the Australian standard.

@@

```{r}
# Read in shape information
tas_shp <- read_shape(shp_path = shp_path)

# Convert to a simple features object
tas_lga <- sf::st_as_sf(tas_shp)

# Select only desired LGA areas, those from Tasmania.
tas_lga <- tas_shp %>% filter(STE_NAME11 == "Tasmania")
tas_lga[[sf_id]] <- droplevels(as.factor(tas_lga[[sf_id]]))

## Not necessary within `create_hexmap`:
## Derive projection information
crs_info <- sf::st_crs(tas_lga)
proj4string <- crs_info$proj4string
epsg <- crs_info$epsg

@@

tas_lga <- tas_lga %>%
    sf::st_transform(., crs = projstring)

## Use the create_centroids function to find polygon centroids:
centroids <- create_centroids(shp_sf = tas_lga, sf_id = sf_id, projstring = projstring, verbose = FALSE)
```

```{r, fig.cap="Tasmanian Local Government Areas and derived centroids."}
fort_tas <- fortify_sfc(tas_lga, projstring = projstring)
ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = group), data = fort_tas, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude), data= centroids, colour = "#b2df8a") + theme_void() + coord_equal()
```

@@

```{r}
## Not necessary within `create_hexmap`:
## create the bounding box from the set of polygon centroids:
bbox <- tibble::tibble(min = c(min(centroids$longitude), min(centroids$latitude)),
        max = c(max(centroids$longitude), max(centroids$latitude)))

hex_size <- (bbox$max[1] - bbox$min[1])/(bbox$max[2] - bbox$min[2]) / 10
hex_size <- 0.2

@@

```{r, fig}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_tas, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude), data= centroids, colour = "#b2df8a") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75) 
```

@@

```{r}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_tas, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroids, fill = "#b2df8a", fill = "red") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75) 
```

@@

```{r}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_tas, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroids, fill = "#b2df8a", fill = "red") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = full_grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75)
```

@@

```{r}
## Not necessary within `create_hexmap`:
## Consider hexagon points up to a certain distance from centroid

@@

Create a box filter
```{r, echo = FALSE}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_tas, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroid1, fill = "red") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75)

@@

```{r, echo = FALSE}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_tas, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroid1, fill = "red") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75)
```

@@

ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_tas, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroid1, fill = "red") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75)

@@

```{r}
allocation <- centroid_allocation
hexmap_lga <- left_join(tas_lga, allocation)
ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = group), data = fort_tas, fill = "grey", colour = "white") +
  theme_void() + coord_equal() +
    geom_hex(aes(x = hex_long, y = hex_lat, fill = AREA_SQKM), data = hexmap_lga, 
      position = "identity", stat = "identity")
```

```{r, eval = FALSE}
hexmap_lga <- left_join(tas_lga, 
  create_hexmap(shp = tas_lga, sf_id = "LGA_CODE11", buffer_dist = NULL, filter_dist = 10, hex_size = 0.2, export_shp = FALSE, focal_points = sugaRbag::capital_citie, verbose = TRUE)
)
```

@@

Examples of Melbourne, Tasmania, Australia
At different levels, distance from original centroids, change in area

LGA of all Australia
SA2 for Queensland

>>>>

---
title: "An algorithm for a spatial mapping using a hexagon tilegram, with application to Australia disease maps"
author:
  - name: Stephanie Kobakian
    url: https://github.com/srkobakian
    affiliation: nswmania University of Technology
date: "`r Sys.Date()`"
output: radix::radix_article
---

@@

Spatial distrbutions have utilised alternative representations of geography for many years. In modern times with the ease of interactivity and animation, alternative representations have been popularised by online news sites. Applications are iincreasingly widespread, especially in the area of disease mapping.

@@

Spatial distributions of people and human related statistics are not well presented in geographic maps. Populations congregate around major cities and the lack of land available encourages vertical living. Government bodies define localities, attempting to evenly distribute the population into measureable areas. As the rural areas of Australia are often sparsely populated in comparison to their urban centres the square meterage of the geographic areas may vastly differ in size.

Alternative mapping methods allow increased understanding of the spatial distribution of a variable amongst the population. Allowing the focus to be placed on the distribution of the statistics between the groups of areas, not the geography.

@@

Extremely small geographic areas are lost on large geographic maps, areas in Sydney or Melbourne are not easily compared at a high level.
Cartograms allow an alternative view. Placingimportance on the statistic of interest, allowing unequal map space on the display to represent differences in the statistic. These maps rely on the statistic of interest to determine the layout, and often fail to preserve a recognisable view of Australia.
Tile grams allow a better preservation of relationships. Decrease the emphasis on the amount of geograhic area considered to be interesting. Focus on the relationship between neighbours, in considering aggregated statistics of heterogeneous regions.

@@

Australia and Canada present an application that contains an unusual population settlement and dispersion over the geographical space: 
- Australia: Dense south east, sparse central and west

This provides an opportunity for a new class of

@@

The algorithm operates on a set of polygons.
There are parameters used in the process that may be provided, or will be automatically derived.
All necessary functions are exported, with a main function `create_hexmap` used to step through automatically.

@@

As the `create_hexmap` function requires several parameters, if they are not provided, the information will be derived from the polygon set used.

**The following must be provided to `create_hexmap`:**
```{r table_necessary}
knitr::kable(tibble::tibble(parameter = c('shp', 
  'shp_path',
  'sf_id',
  'focal_points'),
  description = c('a shape file', 'character string location of shape file', 'name of a unique column that distinguishes areas',
    'a data frame of reference locations when allocating hexagons, capital cities of Australia are provided as capital_cities'
    )))

```{r data}
library(tidyverse)
library(sugaRbag)

@@

**The following will be determined within `create_hexmap` if not provided. They are created throughout the following example:**

```{r}
knitr::kable(tibble(parameter = c('projstring', 
  'epsg', 'buffer_dist', 'hex_size',  'filter_dist', 'width'),
  description = c('a string to indicate the projection and epsg', 'the four character string to indicate the CRS', 
    'a float value for distance in degrees to extend beyond the - geometry provided',
    'a float value in degrees for the diameter of the hexagons', 'amount of hexagons around centroid to consider for allocation',
    'the angle used to filter the grid points around a centroid')))
```

@@

When utilising individual `sugaRbag` functions, we recommend the following approach:

@@

Begin with a set of polygons.
There are several ways of providing the set of polygons.
The following methods are available via the `read_shape` function. It utilises the `st_read` function is to create the sf object.

- To utilise a shape file from a directory on your device, provide the file path for .shp file.
- If the shape file is an .Rda file, it is loaded into the current environment.

Empty geometries are removed from the sf object, and the remaining polygons are projected to the Australian standard using `projstring`.

@@

```{r}
# Read in shape information
nsw_shp <- read_shape(shp_path = shp_path)

# Convert to a simple features object
nsw_lga <- sf::st_as_sf(nsw_shp)

# Select only desired LGA areas, those from nswmania.
nsw_lga <- nsw_shp %>% filter(STE_NAME11 == "New South Wales")
nsw_lga[[sf_id]] <- droplevels(as.factor(nsw_lga[[sf_id]]))

## Not necessary when using `create_hexmap`:
## Derive projection information
crs_info <- sf::st_crs(nsw_lga)
proj4string <- crs_info$proj4string
epsg <- crs_info$epsg

@@

nsw_lga <- nsw_lga %>%
    sf::st_transform(., crs = projstring)

## Use the create_centroids function to find polygon centroids:
centroids <- create_centroids(shp_sf = nsw_lga, sf_id = sf_id, projstring = projstring, verbose = FALSE)
```

```{r, fig.cap="nswmania Local Government Areas and derived centroids."}
fort_nsw <- fortify_sfc(nsw_lga, projstring = projstring)
ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = group), data = fort_nsw, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude), data= centroids, colour = "#b2df8a") + theme_void() + coord_equal()
```

@@

```{r}
## Not necessary when using `create_hexmap`:
## create the bounding box from the set of polygon centroids:
bbox <- tibble::tibble(min = c(min(centroids$longitude), min(centroids$latitude)),
        max = c(max(centroids$longitude), max(centroids$latitude)))

hex_size <- (bbox$max[1] - bbox$min[1])/(bbox$max[2] - bbox$min[2]) / 5

@@

```{r}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_nsw, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude), data= centroids, colour = "#b2df8a") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75) 
```

@@

```{r}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_nsw, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroids, fill = "#b2df8a") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75) 
```

@@

```{r}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_nsw, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroids, fill = "#b2df8a", fill = "red") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = full_grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = grid, size = 0.75)
```

@@

```{r}
## Not necessary when using `create_hexmap`:
## Consider hexagon points up to a certain distance from centroid

@@

Create a box filter
```{r, echo = FALSE}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_nsw, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroid1, fill = "red") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75)

@@

```{r, echo = FALSE}
ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_nsw, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroid1, fill = "red") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75)
```

@@

ggplot() + geom_polygon(aes(x = long, y = lat, group = group), data = fort_nsw, fill = "grey", colour = "white") + geom_point(aes(x=longitude, y = latitude),data= centroid1, fill = "red") + theme_void() + coord_equal() +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#a6cee3", data = grid, size = 0.25) +
  geom_point(aes(x = hex_long, y = hex_lat), colour = "#1f78b4", data = hex_grid, size = 0.75)

@@

```{r fullhexmap}
lga_map <- create_hexmap(shp = nsw_lga, sf_id = "LGA_CODE11", buffer_dist = NULL, filter_dist = 10, export_shp = FALSE, focal_points = sugaRbag::capital_cities, projstring = projstring, verbose = TRUE)

hexmap_lga <- left_join(nsw_lga, 
  lga_map)

```{r}
ggplot() + 
  geom_polygon(aes(x = long, y = lat, group = group), data = fort_nsw, fill = "grey", colour = "white") +
  theme_void() + coord_equal() +
    geom_hex(aes(x = hex_long, y = hex_lat, fill = AREA_SQKM), data = hexmap_lga, 
      position = "identity", stat = "identity")
```

@@

Examples of Melbourne, nswmania, Australia
At different levels, distance from original centroids, change in area

LGA of all Australia
SA2 for nswmania
